---
title: "Understanding Split-Plot design: A study notebook"
author: "Andrea Bazzano"
format: html
execute: 
  echo: true
  warning: false
  message: false
---

```{r}
#| label: libraries_functions

library(knitr)
source("R/create_dataset.R")
```

## Why This Example Stopped Me

Table 9.1 from Box, Hunter & Hunter, presents a study case where the effect of temperature and coating on corrosion resistance has to be studied.

A furnace is set at three different temperatures twice and for each heat setting four differently coated bars were randomly arranged in the furnace.

The authors highlight that setting the furnace temperature accurately was difficult.

I initially thought:

>  This is just a two-factor experiment. Temperature × Coating.

But something felt wrong.

The model ran in R.
The ANOVA table looked clean but I could not reproduce the results in the book.

This document is my attempt to understand:

- What the actual experimental units are
- Why this is not a completely randomized design
- What changes in the ANOVA table
- What I learned by reconstructing it in R

## Reconstructing the Dataset

The experiment studies corrosion resistance of steel bars:

- Temperature (Temp): 360, 370, 380 °C
- Coating (Coat): C1–C4
- Each temperature applied to a furnace run (Heat)
- Within each heat, four coated bars are tested.

```{r}
#| label: tbl-dataset
#| tbl-cap: "Data from table 9.1 in Box, Hunter & Hunter - Statistics for Experimenters, (2005), Design, Innovation and Discovery, Second Edition. `y` is corrosion resistance: the higher, the better."

dataset <- create_dataset()
knitr::kable(dataset)
```

At first glance, this looks like a balanced design with three levels for temperature and four for coatings, replicated twice.

But it is not.

## If This Were a Completely Randomized Design

Suppose all 24 observations were independent experimental units.

Then the model would be:

```{r}
#| label: crd_design

model_crd <- aov(y ~ Temp * Coat, data = dataset)
anova(model_crd)
```

This has a great impact on the actual experimental plan:

> The furnace temperature now has to be set 24 times at random, for only a differently coated bar for each run: a very inefficient setup resulting in widely varying estimates because of the difficulty associated to setting the furnace temperature.

With this design:

- Each bar is independently randomized to temperature and coating.
- There is one common residual error term.

But is that true?

No. The experimental plan described in the textbook is very different.

- Temperature was applied to the entire furnace run.
- Each coating was tested within each furnace run.
- The furnace run — not the bar — is the experimental unit for temperature.

Already, something is structurally wrong.

## Identifying the Experimental Units

To clarify, I forced myself to distinguish:

**Unit of observation** → Individual steel bar

**Unit of randomization for Temperature** → Furnace run (Heat)

**Unit of randomization for Coating** → Bar within Heat

This reveals the hierarchy:

```
Heat
└── 4 bars (different coatings)
```

There are two levels of randomization.

Therefore, there must be two error terms.

This is the defining feature of a split-plot design.

## Fitting the Split-Plot Model

In base R, the classical way to represent this structure is:

```{r}
#| label: split_plot

model_sp <- aov(y ~ Temp * Coat + Error(Heat), data = dataset)
summary(model_sp)
```


R now separates the analysis into strata:

- Heat stratum (**whole-plot** error)
- Within stratum (**subplot** error)

Conceptually:

- Temperature is tested against variability between heats
- Coating and Temperature × Coating  are tested against variability within heats
- This changes the denominator of the F-tests.
- And that changes the conclusions.

The variability introduced by setting the furnace temperature is not eliminated.
It is acknowledged as a separate source of experimental error.

## Comparing the Two Analyses

Let’s compare directly the two designs.

```{r}
#| label: comparison

summary(model_crd)
summary(model_sp)
```


In the completely randomized model:

Temperature is tested against the residual mean square
→ Treating 24 bars as independent for temperature

In the split-plot model:

Temperature is tested against mean square of Heat
→ Correctly using only between-heat variability

The difference reflects the true replication structure.

I had 8 observations per temperature level, but only 2 independent experimental units.

That realization was the turning point.

## Degrees of Freedom Logic

Let me write explicitly what the structure implies.

- Temperatures: 3 levels
- Heats: 6 total (2 per temperature)
- Coatings: 4 levels
- 24 observations, then 23 total degrees of freedom (df)

**Whole-plot level**:

  - the number of temperature levels subtracted the grand average gives 3 − 1 = 2 df
  - the number of heats minus the number of temperature levels gives 6 − 3 = 3 df, for testing the temperature effect.

**Subplot level**:

  - the number of coatings gives 4 − 1 = 3 df
  - the interaction between temperature and coating gives (3 − 1)(4 − 1) = 2 × 3 = 6 df
  - the total df subtracted of the df for the other terms gives 23 − 14 = 9 df for testing the coating effect and its interaction with temperature.

The CRD model ignores this decomposition and gives:

- Intercept: 1 df
- Temperature: 2 df
- Coating: 3 df
- Interaction: 6 df
- 24 − 12 = 12 df for testing all the effects.

It borrows precision from the subplots when testing temperature.
That artificially inflates significance.

The completely randomized model gives 12 residual degrees of freedom.
The split-plot model gives only 9.
The missing 3 degrees of freedom correspond exactly to whole-plot variation.

## Why This Is Not a Randomized Block Design

If temperature were a block, this would mean:

- Temperature is not of primary interest and it only serves to remove variability.
- Coatings would be compared within temperature.

But here:

- Temperature is a treatment factor.
- Each heat receives only one temperature.
- Heats are the replicates of temperature.
- Heat is not blocking temperature.
- Heat is the experimental unit for temperature.

Confusing these roles changes the denominator of the F-test.

As the authors stated:

> The split-plot is like a randomized block design (with whole plots as blocks) in which the opportunity is taken to introduce additional factors *between blocks*.

A sentence that gave me a lot to think!

## What I Learned

ANOVA is not just mechanics.
It is inseparable from the data model and the experimental plan.
That is the source of both its complexity and its flexibility.

Well done, R. A. Fisher.

- The experimental unit determines the correct error term.
- Split-plot designs arise from constrained randomization.
- Ignoring hierarchy produces incorrect F-tests.
- What looks like a simple factorial can hide multiple sources of variation.

Most importantly:

I had been testing temperature against the wrong error term
without realizing it.

That is not a computational mistake.
It is a conceptual one.

## Next Step

This classical aov() formulation implicitly reflects a mixed model:

- Temp fixed
- Heat random
- Coat fixed within Heat

Understanding this connection is the next conceptual step.

For now, this notebook clarified something essential:

- Split-plot is not a technical complication.
- It is the statistical consequence of how the experiment was physically run.